\textbf{Seasonal Dynamics of Epiphytic Microbial Communities on Marine Macrophyte Surfaces}

### 2020-05-28 submission/manuscript.tex
Review optimization.

3. Department of Marine Microbiology and Biogeochemistry, Royal
Netherlands Institute for Sea Research, Utrecht University, The
Netherlands

@@

Surfaces of marine macrophytes (seagrasses and macroalgae) are inhabited
by diverse microbial communities. Most studies focusing on macrophyte
epiphytic communities did not take into account temporal changes or
applied low sampling frequency approaches. Illumina sequencing of the V4
16S rRNA region was performed to determine the seasonal dynamics of
epiphytic communities sampled from the surfaces of the seagrass
\emph{Cymodocea nodosa} and invasive macroalga \emph{Caulerpa
cylindracea}. Leaves and thalli were sampled in a meadow of
\emph{Cymodocea nodosa} invaded by the invasive \emph{Caulerpa
cylindracea} and in a monospecific settlement of \emph{Caulerpa
cylindracea} located in the proximity of the meadow at monthly
intervals. For comparison the ambient prokaryotic plankton community was
also characterized. Sequencing results at the OTU level showed a clear
differentiation between ambient water and epiphytic communities and a
host-specific community assemblage. In addition, successional changes
were observed that could be connected to the macrophyte growth cycle.
Taxonomic analysis showed similar high rank groups in the ambient water
and epiphytic communities, with the exception of \emph{Desulfobacterota}
that were found only on \emph{Caulerpa cylindracea}. Only
\emph{Cyanobacteria} showed seasonal change, while other high rank taxa
were present throughout the year. In every analyzed high rank taxa,
phylogenetic groups present throughout the year comprised most of the
sequences and could be identified together with low proportion taxa
showing seasonal patterns connected to the macrophyte growth cycle.
Taken together, epiphytic microbial communities of the seagrass
\emph{Cymodocea nodosa} and the macroalgae \emph{Caulerpa cylindracea}
appear to be host-specific and contain taxa that undergo successional
changes.

@@

Marine macrophytes (seagrasses and macroalgae) are important ecosystem
engineers that form close associations with microorganism belonging to
all three domains of life (Egan \emph{et al.}, 2013; Tarquinio \emph{et
al.}, 2019). Microbes can live within macrophyte tissue as endophytes or
can form epiphytic communities on surfaces of leaves and thalli (Egan
\emph{et al.}, 2013; Hollants \emph{et al.}, 2013; Tarquinio \emph{et
al.}, 2019). Epiphytic and endophytic microbial communities form a close
functional relationship with the macrophyte host. It was proposed that
this close relationship constitutes a holobiont, an integrated community
where the macrophyte organism and its symbiotic partners support each
other (Margulis, 1991; Egan \emph{et al.}, 2013; Tarquinio \emph{et
al.}, 2019).

Biofilms formed from microbial epiphytes can contain diverse taxonomic
groups and harbor cell densities from 10\textsuperscript{2} to
10\textsuperscript{7} \si{\cells\per\cm\squared} (Armstrong \emph{et
al.}, 2000; Bengtsson \emph{et al.}, 2010; Burke and Thomas \emph{et
al.}, 2011). In such an environment a number of positive and negative
interactions between the macrophyte and colonizing microorganisms have
been described (Egan \emph{et al.}, 2013; Hollants \emph{et al.}, 2013;
Tarquinio \emph{et al.}, 2019). Macrophytes can promote growth of
associated microbes by nutrient exudation, while in return
microorganisms may support macrophyte performance through improved
nutrient availability, phytohormone production and protection form toxic
compounds, oxidative stress, biofouling organisms and pathogens (Egan
\emph{et al.}, 2013; Hollants \emph{et al.}, 2013; Tarquinio \emph{et
al.}, 2019). Beside this positive interactions, macrophytes can
negatively impact the associated microbes such as pathogenic bacteria by
producing reactive oxygen species and secondary metabolites (Egan
\emph{et al.}, 2013; Hollants \emph{et al.}, 2013; Tarquinio \emph{et
al.}, 2019).

All these ecological roles are carried out by a taxonomically diverse
community of microorganisms. At higher taxonomic ranks a core set of
macrophyte epiphytic taxa was described consisting mainly of
\emph{Alphaproteobacteria}, \emph{Gammaproteobacteria},
\emph{Desulfobacterota}, \emph{Bacteroidota}, \emph{Cyanobacteria},
\emph{Actinobacteriota}, \emph{Firmicutes}, \emph{Planctomycetota},
\emph{Chloroflexi} and \emph{Verrucomicrobiota} (Crump and Koch, 2008;
Tujula \emph{et al.}, 2010; Lachnit \emph{et al.}, 2011). In contrast,
at lower taxonomic ranks host specific microbial communities were
described (Lachnit \emph{et al.}, 2011; Roth-Schulze \emph{et al.},
2016). Recently, it was shown that even different morphological niches
within the same alga had a higher influence on bacterial community
variation than biogeography or environmental factors (Morrissey \emph{et
al.}, 2019). While there is high community variation between host
species it was observed that the majority of metagenome determined
functions were conserved both between host species and individuals
(Burke and Peter Steinberg \emph{et al.}, 2011; Roth-Schulze \emph{et
al.}, 2016). This discrepancy between taxonomic and functional
composition could be explained by the lottery hypothesis. It postulates
that an initial random colonization step is performed from a set of
functionally equivalent taxonomic groups resulting in taxonomically
different epiphytic communities sharing a core set of functional genes
(Burke and Peter Steinberg \emph{et al.}, 2011; Roth-Schulze \emph{et
al.}, 2016). In addition, some of the variation in the observed data
could be attributed to different techniques used in various studies,
such as different protocols for epiphytic cell detachment and/or DNA
isolation, as no standard protocol to study epiphytic communities was
established (Ugarelli \emph{et al.}, 2019; Korlević \emph{et al.},
submitted).

The majority of studies describing macrophyte epiphytic communities did
not encompass seasonal changes (Crump and Koch, 2008; Lachnit \emph{et
al.}, 2009; Burke and Thomas \emph{et al.}, 2011; Roth-Schulze \emph{et
al.}, 2016; Ugarelli \emph{et al.}, 2019). In addition, if seasonal
changes were taken into account low temporal frequency and/or
methodologies that do not allow high taxonomic resolution were used
(Tujula \emph{et al.}, 2010; Lachnit \emph{et al.}, 2011; Miranda
\emph{et al.}, 2013; Michelou \emph{et al.}, 2013). In the present study
we describe the seasonal dynamics of bacterial and archaeal communities
on the surfaces of the seagrass \emph{Cymodocea nodosa} and siphonous
macroalgae \emph{Caulerpa cylindracea} determined on a mostly monthly
scale. Bacterial and archaeal epiphytes were sampled in a meadow of
\emph{C. nodosa} invaded by the invasive \emph{C. cylindracea} and in a
locality of only \emph{C. cylindracea} located in the proximity of the
meadow. In addition, for comparison, the community of the ambient
seawater was characterized.

@@

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of eukaryotic,
chloroplast, mitochondrial and no relative sequences
(\autoref{supp-nseq_notus}). A total of 35 samples originating from
epiphytic archaeal and bacterial communities associated with surfaces of
the seagrass \emph{C. nodosa} and macroalga \emph{C. cylindracea} were
analyzed. In addition, 18 samples (one of the samples was sequenced two
times) originating from picoplankton archaeal and bacterial communities
in the ambient seawater were also processed for comparison. The number
of reads per sample ranged between 8,408 and 77,463 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off that is a common
observation in high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences mentioned before reads were clustered into 28,750 different
OTUs at a similarity level of 97 \si{\percent}. Read numbers were
normalized to the minimum number of sequences, 8,408
(\autoref{supp-nseq_notus}), through rarefaction resulting in 17,017
different OTUs that ranged from 360 to 2,108 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes of richness
and diversity the Observed Number of OTUs, Chao1, ACE, Exponential
Shannon (Jost, 2006) and Inverse Simpson were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{C. cylindracea} (mixed {[}Number of OTUs, 1,675.9 ±
144.0 OTUs{]} and monospecific {[}Number of OTUs, 1,735.6 ± 175.0
OTUs{]}), middle values for \emph{C. nodosa} (Number of OTUs, 1,052.6 ±
204.9 OTUs) and lower values for picoplankton communities in the ambient
seawater (Number of OTUs, 525.3 ± 146.6 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not show such large
dissimilarities. \emph{C. nodosa} communities showed a slow increase
towards the end of the study, while \emph{C. cylindracea} (mixed and
monospecific) communities were characterized by slightly larger values
in Spring and Summer in comparison to Autumn and Winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient were, respectively, calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from a
particular environment. The highest proportion of shared OTUs and
community was found between mixed and monospecific \emph{C. cylindracea}
(Jaccard, 0.36; Bray-Curtis, 0.78), while lower shared values were
calculated between seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{C. nodosa} and
\emph{C. cylindracea} were approximately in the middle between these two
extremes. To assess seasonal changes in the proportion of shared OTUs
and communities the Jaccard's and Bray-Curtis Similarity Coefficients
were calculated between consecutive sampling points (\autoref{shared}).
Both coefficients showed similar trends. Temporal proportional changes
were more pronounced for seawater in comparison to \emph{C. nodosa} and
especially \emph{C. cylindracea} associated communities
(\autoref{shared}). In addition, only 0.4 -- 1.0 \si{\percent} of OTUs
from each surface associated community were found at every time point.
These OTUs also made a high proportion of total sequences (40.6 -- 52.6
\si{\percent}). To further disentangle the environmental and seasonal
community dissimilarity a Principal Coordinates Analysis (PCoA) based on
Bray-Curtis distances and OTU abundances was applied. It showed a clear
separation between planktonic and surface associated communities
(\autoref{pcoa}). In addition, a separation of epiphytic bacterial and
archaeal communities based on host species was determined. This
separation was further supported by ANOSIM (R = 0.96, \emph{p}
\textless{} 0.001). Seasonal changes of \emph{C. nodosa} associated
communities indicated a separation between Spring, Summer and
Autumn/Winter samples (ANOSIM, R = 0.53, \emph{p} \textless{} 0.01). For
\emph{C. cylindracea} associated communities a separation between Summer
and Autumn/Winter/Spring samples was observed that was not so strongly
supported (ANOSIM, R = 0.31, \emph{p} \textless{} 0.05)
(\autoref{pcoa}).

The taxonomic composition of both, macrophyte associated and seawater
communities, was dominated by bacterial (99.1 ± 2.1 \si{\percent}) over
archaeal sequences (0.9 ± 2.1 \si{\percent}) (\autoref{community}).
Higher relative abundances of chloroplast related sequences were only
observed in surface associated communities, with higher values in
Autumn/Winter (37.2 ± 11.2 \si{\percent}) in comparison to Spring/Summer
(20.9 ± 9.7 \si{\percent}) (\autoref{supp-chloroplast}). Generally, at
higher taxonomic ranks (phylum-class) epiphytic and seawater microbial
communities were composed of similar bacterial taxa. Seawater
communities were mainly comprised of \emph{Actinobacteriota},
\emph{Bacteroidota}, \emph{Cyanobacteria}, \emph{Alphaproteobacteria},
\emph{Gammaproteobacteria} and \emph{Verrucomicrobiota}. Communities
associated with \emph{C. nodosa} consisted of same groups with the
addition of \emph{Planctomycetota} whose contribution was higher in
summer 2018. In addition, communities from mixed and monospecific
\emph{C. cylindracea} were similar and characterized by same groups as
seawater and \emph{C. nodosa} communities with the addition of
\emph{Desulfobacterota} (\autoref{community}). Larger differences
between environments and host species could be observed at lower
taxonomic ranks (\autoref{cyano} -- \ref{desulfo}).

\emph{Cyanobacteria} related sequences were comprising, on average, 5.5
± 4.4 \si{\percent} of total sequences (\autoref{cyano}). Higher
proportions were found for \emph{C. nodosa} (16.4 ± 5.3 \si{\percent})
and \emph{C. cylindracea} (mixed {[} (7.7 ± 3.9 \si{\percent}){]} and
monospecific {[} (7.8 ± 2.4 \si{\percent}){]}) associated communities in
autumn and for seawater communities in winter (8.8 ± 7.5 \si{\percent}).
Large taxonomic differences between surface associated and seawater
cyanobacterial communities were observed. Seawater communities were
mainly comprised of \emph{Cyanobium} and \emph{Synechococcus}, while
surface associated communities were consisted of \emph{Pleurocapsa} and
sequences without known relatives within \emph{Cyanobacteriia}
(\autoref{cyano}). In addition, seasonal changes in surface associated
communities were observed with \emph{Pleurocapsa} and no relative
\emph{Cyanobacteriia} comprising larger proportions in autumn and winter
and \emph{Acrophormium}, \emph{Phormidesmis} and no relative
\emph{Nodosilineaceae} in spring and summer (\autoref{cyano}).

Sequences classified as \emph{Bacteroidota} were comprising, on average,
19.2 ± 5.5 \si{\percent} of all sequences (\autoref{bactero}). Similarly
to \emph{Cyanobacteria}, large differences in the taxonomic composition
between seawater and surface associated communities were found
(\autoref{bactero}). The seawater community was characterized by the NS4
and NS5 marine groups, uncultured \emph{Cryomorphaceae}, uncultured
\emph{Flavobacteriaceae}, NS11-12 marine group, \emph{Balneola},
uncultured \emph{Balneolaceae} and \emph{Formosa}. In contrast, in
surface associated communities \emph{Lewinella}, \emph{Portibacter},
\emph{Rubidimonas}, no relative \emph{Saprospiraceae}, uncultured
\emph{Saprospiraceae}, no relative \emph{Flavobacteriaceae} and
uncultured \emph{Rhodothermaceae} were found. Some groups showed slight
seasonal changes such as no relative \emph{Flavobacteriaceae} that were
more pronounced from November 2017 until June 2018. In contrast,
uncultured \emph{Rhodothermaceae} showed higher proportions from June
2018 until the end of the study period. Surface associated
\emph{Bacteroidota} communities were very diverse as could be observed
in the high proportion of taxa that grouped as other \emph{Bacteroidota}
(\autoref{bactero}).

On average, \emph{Alphaproteobacteria} were in comparison to other high
rank taxa the largest taxonomic group, comprising 29.2 ± 12.0
\si{\percent} of all sequences (\autoref{alpha}). In accordance to
previous taxa, high differences between seawater and surface associated
communities were observed. Picoplankton communities were composed mainly
of the SAR11 clade, AEGEAN-169 marine group, SAR116 clade, no relative
\emph{Rhodobacteraceae}, HIMB11 and OCS116 clade, while surface
associated communities were composed in high proportion of no relative
\emph{Rhodobacteraceae} and to a lesser degree of \emph{Pseudoahrensia},
no relative \emph{Alphaproteobacteria}, no relative
\emph{Hyphomonadaceae} and \emph{Amylibacter}. Representatives of no
relative \emph{Rhodobacteraceae} were comprising on average 40.6 ± 23.2
\si{\percent} of all alphaproteobacterial sequences from the epiphytic
community (\autoref{alpha}). In addition, \emph{Amylibacter} was
detected mainly in \emph{C. nodosa} from November 2017 until March 2018.

Sequences related to \emph{Gammaproteobacteria} were comprising, on
average, 18.6 ± 3.9 \si{\percent} of all sequences (\autoref{gamma}).
Similarly to previous taxa, large taxonomic differences between seawater
and surface associated communities were found. Seawater communities were
mainly comprised of the OM60 (NOR5) clade, \emph{Litoricola},
\emph{Acinetobacter} and the SAR86 clade, while epiphytic communities
were mainly composed of no relative \emph{Gammaproteobacteria} and
\emph{Granulosicoccus}. Beside these two groups specific to all three
epiphytic communities, \emph{C. nodosa} was characterized by
\emph{Arenicella}, no relative \emph{Burkholderiales} and
\emph{Methylotenera}, while \emph{Thioploca}, no relative
\emph{Cellvibrionaceae} and \emph{Reinekea} were more specific to both
mixed and monospecific \emph{C. cylindracea}. In addition,
\emph{Arenicella} was more pronounced in November and December 2017,
while no relative \emph{Burkholderiales} and \emph{Methylotenera} were
more characteristic for the period form March until May 2018. For the
\emph{C. cylindracea} specific taxa no relative \emph{Cellvibrionaceae}
and \emph{Reinekea} showed some seasonality and were characteristic for
samples originating from June to October 2018. In addition, similarly to
\emph{Bacteroidota}, a large proportion of the surface associated
community was grouped as other \emph{Gammaproteobacteria} indicating
high diversity within this group (\autoref{gamma}).

>>>>

3. NIOZ, Department of Marine Microbiology and Biogeochemistry, Royal
Netherlands Institute for Sea Research, Utrecht University, The
Netherlands

@@

Surfaces of marine macrophytes (seagrasses and macroalgae) are inhabited
by diverse microbial communities. Most studies focusing on epiphytic
communities of macrophytes did not take into account temporal changes or
applied low sampling frequency approaches. Illumina sequencing of the V4
16S rRNA region was performed to determine the seasonal dynamics of
epiphytic communities of the seagrass \emph{Cymodocea nodosa} and
invasive macroalga \emph{Caulerpa cylindracea}. Leaves and thalli were
sampled in a meadow of \emph{Cymodocea nodosa} invaded by the invasive
\emph{Caulerpa cylindracea} and in a monospecific settlement of
\emph{Caulerpa cylindracea} located in the proximity of the meadow at
monthly intervals. For comparison the ambient prokaryotic plankton
community was also characterized. At the OTU level, the microbial
community composition differed between the ambient water and the
epiphytic communities exhibiting host-specificity. Also, successional
changes were observed connected to the macrophyte growth cycle.
Taxonomic analysis, however, showed similar high rank groups in the
ambient water and the epiphytic communities, with the exception of
\emph{Desulfobacterota}, which were only found on \emph{Caulerpa
cylindracea}. \emph{Cyanobacteria} showed seasonal changes while other
high rank taxa were present throughout the year. Phylogenetic groups
present throughout the year constituted most of the sequences, while
less abundant taxa showed seasonal patterns connected to the macrophyte
growth cycle. Taken together, epiphytic microbial communities of the
seagrass \emph{Cymodocea nodosa} and the macroalgae \emph{Caulerpa
cylindracea} appear to be host-specific and contain taxa that undergo
successional changes.

@@

Marine macrophytes (seagrasses and macroalgae) are important ecosystem
engineers forming close associations with microorganism belonging to all
three domains of life (Egan \emph{et al.}, 2013; Tarquinio \emph{et
al.}, 2019). Microbes can live within macrophyte tissue as endophytes or
can form epiphytic communities on surfaces of leaves and thalli (Egan
\emph{et al.}, 2013; Hollants \emph{et al.}, 2013; Tarquinio \emph{et
al.}, 2019). Epiphytic and endophytic microbial communities exhibit a
close functional relationship with the macrophyte host. It was proposed
that this close relationship constitutes a holobiont, an integrated
community where the macrophyte organism and its symbiotic partners
support each other (Margulis, 1991; Egan \emph{et al.}, 2013; Tarquinio
\emph{et al.}, 2019).

Biofilms of microbial epiphytes can contain diverse taxonomic groups and
harbor cell abundances from 10\textsuperscript{2} to
10\textsuperscript{7} \si{\cells\per\cm\squared} (Armstrong \emph{et
al.}, 2000; Bengtsson \emph{et al.}, 2010; Burke and Thomas \emph{et
al.}, 2011). In such an environment a number of positive and negative
interactions between the macrophyte and the colonizing microorganisms
have been described (Egan \emph{et al.}, 2013; Hollants \emph{et al.},
2013; Tarquinio \emph{et al.}, 2019). Macrophytes can promote growth of
associated microbes by nutrient exudation, while in return
microorganisms may support macrophyte performance through improved
nutrient availability, phytohormone production and protection from toxic
compounds, oxidative stress, biofouling organisms and pathogens (Egan
\emph{et al.}, 2013; Hollants \emph{et al.}, 2013; Tarquinio \emph{et
al.}, 2019). Besides these positive interactions, macrophytes can
negatively impact the associated microbes such as pathogenic bacteria by
producing reactive oxygen species and secondary metabolites (Egan
\emph{et al.}, 2013; Hollants \emph{et al.}, 2013; Tarquinio \emph{et
al.}, 2019).

All these ecological roles are carried out by a taxonomically diverse
community of microorganisms. At higher taxonomic ranks a core set of
macrophyte epiphytic taxa was described consisting mainly of
\emph{Alphaproteobacteria}, \emph{Gammaproteobacteria},
\emph{Desulfobacterota}, \emph{Bacteroidota}, \emph{Cyanobacteria},
\emph{Actinobacteriota}, \emph{Firmicutes}, \emph{Planctomycetota},
\emph{Chloroflexi} and \emph{Verrucomicrobiota} (Crump and Koch, 2008;
Tujula \emph{et al.}, 2010; Lachnit \emph{et al.}, 2011). In contrast,
at lower taxonomic ranks host specific microbial communities were found
(Lachnit \emph{et al.}, 2011; Roth-Schulze \emph{et al.}, 2016).
Recently, it was shown that even different morphological niches within
the same alga had a higher influence on the composition of the bacterial
community than biogeography or environmental factors (Morrissey \emph{et
al.}, 2019). While the microbial community composition varies between
host species, metagenomic analyses revealed that the majority of the
microbial functions are conserved (Burke and Peter Steinberg \emph{et
al.}, 2011; Roth-Schulze \emph{et al.}, 2016). This discrepancy between
the microbial taxonomic and functional composition might be explained by
the lottery hypothesis. It postulates that an initial random
colonization step takes places from a set of functionally equivalent
taxonomic groups resulting in taxonomically different epiphytic
communities sharing a core set of functional genes (Burke and Peter
Steinberg \emph{et al.}, 2011; Roth-Schulze \emph{et al.}, 2016). In
addition, some of the variation in the reported data could be attributed
to different techniques used in these studies, such as different
protocols for epiphytic cell detachment and/or DNA isolation, as no
standard protocol has been yet established to study epiphytic
communities (Ugarelli \emph{et al.}, 2019; Korlević \emph{et al.},
submitted).

The majority of studies describing macrophyte epiphytic microbial
communities did not include possible seasonal changes (Crump and Koch,
2008; Lachnit \emph{et al.}, 2009; Burke and Thomas \emph{et al.}, 2011;
Roth-Schulze \emph{et al.}, 2016; Ugarelli \emph{et al.}, 2019). If
seasonal changes were taken into account, low temporal frequency,
applied methodologies and/or limited number of analyzed host species did
not allow a high taxonomic resolution (Tujula \emph{et al.}, 2010;
Lachnit \emph{et al.}, 2011; Miranda \emph{et al.}, 2013; Michelou
\emph{et al.}, 2013; Mancuso \emph{et al.}, 2016). In the present study
we describe the seasonal dynamics of bacterial and archaeal communities
on the surfaces of the seagrass \emph{Cymodocea nodosa} and siphonous
macroalgae \emph{Caulerpa cylindracea} determined on a mostly monthly
scale. Bacterial and archaeal epiphytes were sampled in a meadow of
\emph{C. nodosa} invaded by the invasive \emph{C. cylindracea} and in a
locality of only \emph{C. cylindracea} located in the proximity of the
seagrass meadow. For comparison, the microbial community of the ambient
seawater was also characterized.

@@

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of sequences without
known relatives (no relative sequences) and eukaryotic, chloroplast and
mitochondrial sequences (\autoref{supp-nseq_notus}). A total of 35
samples originating from epiphytic archaeal and bacterial communities
associated with surfaces of the seagrass \emph{C. nodosa} and the
macroalga \emph{C. cylindracea} were analyzed. In addition, 18 samples
(one of the samples was sequenced twice) originating from the ambient
seawater were also processed for comparison. The number of reads per
sample ranged between 8,408 and 77,463 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off as commonly observed in
high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences as mentioned above reads were clustered into 28,750
different OTUs at a similarity level of 97 \si{\percent}. Read numbers
were normalized to the minimum number of sequences (8,408,
\autoref{supp-nseq_notus}) through rarefaction resulting in 17,135
different OTUs with 373 to 2,025 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes in richness
and diversity the observed number of OTUs, Chao1, ACE, Exponential
Shannon (Jost, 2006) and Inverse Simpson were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{C. cylindracea} (mixed {[}Number of OTUs, 1,694.2 ±
133.7 OTUs{]} and monospecific {[}Number of OTUs, 1,737.9 ± 160.5
OTUs{]}) than for \emph{C. nodosa} (Number of OTUs, 1,055.3 ± 209.2
OTUs) and lowest values were obtained for the microbial community of the
ambient seawater (Number of OTUs, 532.6 ± 141.9 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not reveal such large
dissimilarities. \emph{C. nodosa} communities showed a slow increase
towards the end of the study, while \emph{C. cylindracea} (mixed and
monospecific) communities were characterized by slightly higher values
in spring and summer than in autumn and winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient, respectively, were calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from the
particular environment. The highest proportion of shared OTUs and
community was found between mixed and monospecific \emph{C. cylindracea}
(Jaccard, 0.35; Bray-Curtis, 0.77), while lower shared values were
calculated between ambient seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{C. nodosa} and
\emph{C. cylindracea} were approximately in-between the values of mixed
and monospecific \emph{C. cylindracea}. To assess seasonal changes in
the proportion of shared OTUs and communities the Jaccard's and
Bray-Curtis Similarity Coefficients were calculated between consecutive
sampling points (\autoref{shared}). Both coefficients showed similar
trends. Temporal proportional changes were more pronounced for ambient
seawater than for \emph{C. nodosa} and especially \emph{C. cylindracea}
associated communities (\autoref{shared}). In addition, only 0.4 -- 1.0
\si{\percent} of OTUs from each surface associated community were
present at all seasons. These persistent OTUs constituted a high
proportion of total sequences (39.4 -- 51.9 \si{\percent}). To further
disentangle the environmental and seasonal community dissimilarity a
Principal Coordinates Analysis (PCoA) based on Bray-Curtis distances and
OTU abundances was applied. A clear separation between ambient seawater
and surface associated communities was found (\autoref{pcoa}). In
addition, a separation of epiphytic bacterial and archaeal communities
based on host species was detected. This separation was further
supported by ANOSIM (R = 0.96, \emph{p} \textless{} 0.001). Seasonal
changes of \emph{C. nodosa} associated communities indicated a
separation between spring, summer and autumn/winter samples (ANOSIM, R =
0.54, \emph{p} \textless{} 0.01). For \emph{C. cylindracea} associated
communities a separation between summer and autumn/winter/spring samples
was observed that was, however, not as strong as for \emph{C. nodosa}
associated communities (ANOSIM, R = 0.31, \emph{p} \textless{} 0.01)
(\autoref{pcoa}).

The taxonomic composition of both, macrophyte associated and ambient
seawater communities was dominated by bacterial (99.1 ± 2.1
\si{\percent}) over archaeal sequences (0.9 ± 2.1 \si{\percent})
(\autoref{community}). Higher relative abundances of chloroplast related
sequences were only observed in surface associated communities, with
higher values in autumn/winter (37.2 ± 11.2 \si{\percent}) than in
spring/summer (20.9 ± 9.7 \si{\percent}) (\autoref{supp-chloroplast}).
Generally, at higher taxonomic ranks (phylum-class), epiphytic and
ambient seawater microbial communities were composed of similar
bacterial taxa. Ambient seawater communities were mainly comprised of
\emph{Actinobacteriota}, \emph{Bacteroidota}, \emph{Cyanobacteria},
\emph{Alphaproteobacteria}, \emph{Gammaproteobacteria} and
\emph{Verrucomicrobiota}. Communities associated with \emph{C. nodosa}
consisted additionally of \emph{Planctomycetota} contributing more in
summer 2018 than in other seasons. In addition, communities from mixed
and monospecific \emph{C. cylindracea} were similar and characterized by
the same groups as ambient seawater and \emph{C. nodosa} communities
with the addition of \emph{Desulfobacterota} (\autoref{community}).
Larger differences between environments and host species were observed
at lower taxonomic ranks (Figs. \ref{cyano} -- \ref{desulfo}).

\emph{Cyanobacteria} related sequences comprised, on average, 5.5 ± 4.4
\si{\percent} of total sequences (\autoref{cyano}). Higher proportions
were found for \emph{C. nodosa} (16.4 ± 5.3 \si{\percent}) and \emph{C.
cylindracea} mixed (7.7 ± 3.9 \si{\percent}) and monospecific (7.8 ± 2.4
\si{\percent}) associated communities in autumn and for ambient seawater
communities in winter (8.8 ± 7.5 \si{\percent}). Large taxonomic
differences between surface associated and ambient seawater
cyanobacterial communities were observed. Ambient seawater communities
were mainly comprised of \emph{Cyanobium} and \emph{Synechococcus},
while surface associated communities were comprised of
\emph{Pleurocapsa} and sequences within the class \emph{Cyanobacteriia}
that could not be further classified (no relative \emph{Cyanobacteriia})
(\autoref{cyano}). In addition, seasonal changes in surface associated
communities were observed in \emph{Pleurocapsa} and no relative
\emph{Cyanobacteriia} comprising larger proportions in autumn and winter
and \emph{Acrophormium}, \emph{Phormidesmis} and sequences without known
relatives within the \emph{Nodosilineaceae} (no relative
\emph{Nodosilineaceae}) in spring and summer (\autoref{cyano}).

Sequences classified as \emph{Bacteroidota} comprised, on average, 19.2
± 5.5 \si{\percent} of all sequences (\autoref{bactero}). Similarly to
\emph{Cyanobacteria}, large differences in the taxonomic composition
between ambient seawater and surface associated communities were found
(\autoref{bactero}). The ambient seawater community was characterized by
the NS4 and NS5 marine groups, uncultured \emph{Cryomorphaceae},
uncultured \emph{Flavobacteriaceae}, NS11-12 marine group,
\emph{Balneola}, uncultured \emph{Balneolaceae} and \emph{Formosa}. In
contrast, in surface associated communities \emph{Lewinella},
\emph{Portibacter}, \emph{Rubidimonas}, sequences without known
relatives within the \emph{Saprospiraceae} (no relative
\emph{Saprospiraceae}), uncultured \emph{Saprospiraceae}, (sequences
without known relatives within the \emph{Flavobacteriaceae} (no relative
\emph{Flavobacteriaceae})and uncultured \emph{Rhodothermaceae} were
found. Some groups showed minor seasonal changes such as no relative
\emph{Flavobacteriaceae} whose sequences were more abundant from
November 2017 until June 2018. In contrast, uncultured
\emph{Rhodothermaceae} showed higher proportions from June 2018 until
the end of the study period. Surface associated \emph{Bacteroidota}
communities were very diverse as observed in the high proportion of taxa
clustering as other \emph{Bacteroidota} (\autoref{bactero}).

On average, \emph{Alphaproteobacteria} were in comparison to the other
high rank taxa the largest taxonomic group, comprising 29.2 ± 12.0
\si{\percent} of all sequences (\autoref{alpha}). In accordance to the
above described taxa, large differences between ambient seawater and
surface associated communities were observed. Ambient seawater
communities were composed mainly of the SAR11 clade, AEGEAN-169 marine
group, SAR116 clade, sequences without known relatives within the
\emph{Rhodobacteraceae} (no relative \emph{Rhodobacteraceae}), HIMB11
and the OCS116 clade, while surface associated communities were composed
mainly of no relative \emph{Rhodobacteraceae} and to a lesser degree of
\emph{Pseudoahrensia}, \emph{Amylibacter} and sequences without known
relatives within the \emph{Alphaproteobacteria} (no relative
\emph{Alphaproteobacteria}) and \emph{Hyphomonadaceae} (no relative
\emph{Hyphomonadaceae}). Representatives of no relative
\emph{Rhodobacteraceae} comprised on average 40.6 ± 23.2 \si{\percent}
of all alphaproteobacterial sequences in the epiphytic community
(\autoref{alpha}). In addition, \emph{Amylibacter} was detected mainly
in \emph{C. nodosa} from November 2017 until March 2018.

Sequences related to \emph{Gammaproteobacteria} comprised on average
18.6 ± 3.9 \si{\percent} of all sequences (\autoref{gamma}). Similar to
above mentioned taxa, large taxonomic differences between ambient
seawater and surface associated communities were found. Ambient seawater
communities were mainly comprised of the OM60 (NOR5) clade,
\emph{Litoricola}, \emph{Acinetobacter} and the SAR86 clade, while
epiphytic communities were mainly composed of sequences without known
relatives within the \emph{Gammaproteobacteria} (no relative
\emph{Gammaproteobacteria}) and \emph{Granulosicoccus}. Beside these two
groups specific to all three epiphytic communities, \emph{C. nodosa} was
characterized by \emph{Arenicella}, \emph{Methylotenera} and sequences
without known relatives within the \emph{Burkholderiales} (no relative
\emph{Burkholderiales}), while \emph{Thioploca}, \emph{Reinekea} and
sequences without known relatives within \emph{Cellvibrionaceae} (no
relative \emph{Cellvibrionaceae}) were more specific to both mixed and
monospecific \emph{C. cylindracea}. In addition, \emph{Arenicella} was
more pronounced in November and December 2017, while no relative
\emph{Burkholderiales} and \emph{Methylotenera} were characteristic for
the period from March until May 2018. For the \emph{C. cylindracea}
specific taxa no relative \emph{Cellvibrionaceae} and \emph{Reinekea}
showed seasonality and were characteristic for samples originating from
June to October 2018. In addition, similar to \emph{Bacteroidota}, a
large proportion of the surface associated community was grouped as
other \emph{Gammaproteobacteria} indicating high diversity within this
group (\autoref{gamma}).
